---
title: "5. Democratic Transitions"
subtitle: "Comparative Politics"
format: revealjs
---


## From the Last Class

:::{.nonincremental}
- We have seen what a [democracy]{.alert} (early or modern) is

- We have seen how [development]{.alert} may lead to democratic reforms
:::

. . . 

A few issues still remain

- Modern democracy is about [mass]{.alert} participation

- We have assumed that
    - People automatically respond to changing conditions
    - People automatically "act as one" (or coordinate)

- In practice, this may be difficult to achieve


## Our Plan for Today

We will study democratic transitions as [decision-making]{.alert} problems

. . . 

Bottom-up transitions $\rightarrow$ people mobilize for democracy

- Key challenge: mobilizing is risky and people need to do it together

- Two models: [collective action]{.alert} theory and [tipping]{.alert} models

- Explain why revolts are rare and why they surprisingly occur

. . . 

Top-down transitions $\rightarrow$ regime concedes reforms

- Question: why should the regime take the risk?

- Information problems and divisions inside the regime


# Our Case {.section}

[Eastern Europe and the Warsaw Pact]


# Bottom-Up Transitions {.section}

## East Germany 1989


## Collective Action Theory

<table style="margin: 25px auto; font-size: 0.85em; border-collapse: collapse; text-align: center; width: 95%;">
  <thead>
    <tr style="border-bottom: 2px solid #ccc;">
      <th style="padding: 12px 10px; width: 22%;"></th>
      <th style="padding: 12px 10px; width: 26%;">Scenario 1<br><span style="font-size: 0.85em; font-weight: normal;">Fewer than $K-1$ participate</span></th>
      <th style="padding: 12px 10px; width: 26%;">Scenario 2<br><span style="font-size: 0.85em; font-weight: normal;">Exactly $K-1$ participate</span></th>
      <th style="padding: 12px 10px; width: 26%;">Scenario 3<br><span style="font-size: 0.85em; font-weight: normal;">$K$ or more participate</span></th>
    </tr>
  </thead>
  <tbody>
    <tr style="border-bottom: 1px solid #eee;">
      <td style="padding: 16px 10px; font-weight: bold; text-align: right;">Participate</td>
      <td style="padding: 16px 10px;">Pay cost; don't receive benefit</td>
      <td style="padding: 16px 10px;">Pay cost; receive benefit</td>
      <td style="padding: 16px 10px;">Pay cost; receive benefit</td>
    </tr>
    <tr>
      <td style="padding: 16px 10px; font-weight: bold; text-align: right;">Don't participate</td>
      <td style="padding: 16px 10px;">Don't pay cost; don't receive benefit</td>
      <td style="padding: 16px 10px;">Don't pay cost; don't receive benefit</td>
      <td style="padding: 16px 10px;">Don't pay cost; receive benefit</td>
    </tr>
  </tbody>
</table>

---

## Collective Action Theory

<table style="margin: 25px auto; font-size: 0.85em; border-collapse: collapse; text-align: center; width: 95%;">
  <thead>
    <tr style="border-bottom: 2px solid #ccc;">
      <th style="padding: 12px 10px; width: 22%;"></th>
      <th style="padding: 12px 10px; width: 26%; border-top: 3.5px solid #DC143C; border-left: 3.5px solid #DC143C; border-right: 3.5px solid #DC143C; background-color: #FFF5F5;">Scenario 1<br><span style="font-size: 0.85em; font-weight: normal;">Fewer than $K-1$ participate</span></th>
      <th style="padding: 12px 10px; width: 26%;">Scenario 2<br><span style="font-size: 0.85em; font-weight: normal;">Exactly $K-1$ participate</span></th>
      <th style="padding: 12px 10px; width: 26%;">Scenario 3<br><span style="font-size: 0.85em; font-weight: normal;">$K$ or more participate</span></th>
    </tr>
  </thead>
  <tbody>
    <tr style="border-bottom: 1px solid #eee;">
      <td style="padding: 16px 10px; font-weight: bold; text-align: right;">Participate</td>
      <td style="padding: 16px 10px; border-left: 3.5px solid #DC143C; border-right: 3.5px solid #DC143C; background-color: #FFF5F5;">Pay cost; don't receive benefit</td>
      <td style="padding: 16px 10px;">Pay cost; receive benefit</td>
      <td style="padding: 16px 10px;">Pay cost; receive benefit</td>
    </tr>
    <tr>
      <td style="padding: 16px 10px; font-weight: bold; text-align: right;">Don't participate</td>
      <td style="padding: 16px 10px; border-bottom: 3.5px solid #DC143C; border-left: 3.5px solid #DC143C; border-right: 3.5px solid #DC143C; background-color: #FFF5F5;">Don't pay cost; don't receive benefit</td>
      <td style="padding: 16px 10px;">Don't pay cost; don't receive benefit</td>
      <td style="padding: 16px 10px;">Don't pay cost; receive benefit</td>
    </tr>
  </tbody>
</table>

---

## Collective Action Theory

<table style="margin: 25px auto; font-size: 0.85em; border-collapse: collapse; text-align: center; width: 95%;">
  <thead>
    <tr style="border-bottom: 2px solid #ccc;">
      <th style="padding: 12px 10px; width: 22%;"></th>
      <th style="padding: 12px 10px; width: 26%; border-top: 3.5px solid #DC143C; border-left: 3.5px solid #DC143C; border-right: 3.5px solid #DC143C; background-color: #FFF5F5;">Scenario 1<br><span style="font-size: 0.85em; font-weight: normal;">Fewer than $K-1$ participate</span></th>
      <th style="padding: 12px 10px; width: 26%;">Scenario 2<br><span style="font-size: 0.85em; font-weight: normal;">Exactly $K-1$ participate</span></th>
      <th style="padding: 12px 10px; width: 26%;">Scenario 3<br><span style="font-size: 0.85em; font-weight: normal;">$K$ or more participate</span></th>
    </tr>
  </thead>
  <tbody>
    <tr style="border-bottom: 1px solid #eee;">
      <td style="padding: 16px 10px; font-weight: bold; text-align: right;">Participate</td>
      <td style="padding: 16px 10px; border-left: 3.5px solid #DC143C; border-right: 3.5px solid #DC143C; background-color: #FFF5F5;">Pay cost; don't receive benefit</td>
      <td style="padding: 16px 10px;">Pay cost; receive benefit</td>
      <td style="padding: 16px 10px;">Pay cost; receive benefit</td>
    </tr>
    <tr>
      <td style="padding: 16px 10px; font-weight: bold; text-align: right;">Don't participate</td>
      <td style="padding: 16px 10px; border-bottom: 3.5px solid #DC143C; border-left: 3.5px solid #DC143C; border-right: 3.5px solid #DC143C; background-color: #90EE90; font-weight: bold;">Don't pay cost; don't receive benefit</td>
      <td style="padding: 16px 10px;">Don't pay cost; don't receive benefit</td>
      <td style="padding: 16px 10px;">Don't pay cost; receive benefit</td>
    </tr>
  </tbody>
</table>

---

## Collective Action Theory

<table style="margin: 25px auto; font-size: 0.85em; border-collapse: collapse; text-align: center; width: 95%;">
  <thead>
    <tr style="border-bottom: 2px solid #ccc;">
      <th style="padding: 12px 10px; width: 22%;"></th>
      <th style="padding: 12px 10px; width: 26%;">Scenario 1<br><span style="font-size: 0.85em; font-weight: normal;">Fewer than $K-1$ participate</span></th>
      <th style="padding: 12px 10px; width: 26%; border-top: 3.5px solid #DC143C; border-left: 3.5px solid #DC143C; border-right: 3.5px solid #DC143C; background-color: #FFF5F5;">Scenario 2<br><span style="font-size: 0.85em; font-weight: normal;">Exactly $K-1$ participate</span></th>
      <th style="padding: 12px 10px; width: 26%;">Scenario 3<br><span style="font-size: 0.85em; font-weight: normal;">$K$ or more participate</span></th>
    </tr>
  </thead>
  <tbody>
    <tr style="border-bottom: 1px solid #eee;">
      <td style="padding: 16px 10px; font-weight: bold; text-align: right;">Participate</td>
      <td style="padding: 16px 10px;">Pay cost; don't receive benefit</td>
      <td style="padding: 16px 10px; border-left: 3.5px solid #DC143C; border-right: 3.5px solid #DC143C; background-color: #FFF5F5;">Pay cost; receive benefit</td>
      <td style="padding: 16px 10px;">Pay cost; receive benefit</td>
    </tr>
    <tr>
      <td style="padding: 16px 10px; font-weight: bold; text-align: right;">Don't participate</td>
      <td style="padding: 16px 10px; background-color: #90EE90; font-weight: bold;">Don't pay cost; don't receive benefit</td>
      <td style="padding: 16px 10px; border-bottom: 3.5px solid #DC143C; border-left: 3.5px solid #DC143C; border-right: 3.5px solid #DC143C; background-color: #FFF5F5;">Don't pay cost; don't receive benefit</td>
      <td style="padding: 16px 10px;">Don't pay cost; receive benefit</td>
    </tr>
  </tbody>
</table>

---

## Collective Action Theory

<table style="margin: 25px auto; font-size: 0.85em; border-collapse: collapse; text-align: center; width: 95%;">
  <thead>
    <tr style="border-bottom: 2px solid #ccc;">
      <th style="padding: 12px 10px; width: 22%;"></th>
      <th style="padding: 12px 10px; width: 26%;">Scenario 1<br><span style="font-size: 0.85em; font-weight: normal;">Fewer than $K-1$ participate</span></th>
      <th style="padding: 12px 10px; width: 26%; border-top: 3.5px solid #DC143C; border-left: 3.5px solid #DC143C; border-right: 3.5px solid #DC143C; background-color: #FFF5F5;">Scenario 2<br><span style="font-size: 0.85em; font-weight: normal;">Exactly $K-1$ participate</span></th>
      <th style="padding: 12px 10px; width: 26%;">Scenario 3<br><span style="font-size: 0.85em; font-weight: normal;">$K$ or more participate</span></th>
    </tr>
  </thead>
  <tbody>
    <tr style="border-bottom: 1px solid #eee;">
      <td style="padding: 16px 10px; font-weight: bold; text-align: right;">Participate</td>
      <td style="padding: 16px 10px;">Pay cost; don't receive benefit</td>
      <td style="padding: 16px 10px; border-left: 3.5px solid #DC143C; border-right: 3.5px solid #DC143C; background-color: #90EE90; font-weight: bold;">Pay cost; receive benefit</td>
      <td style="padding: 16px 10px;">Pay cost; receive benefit</td>
    </tr>
    <tr>
      <td style="padding: 16px 10px; font-weight: bold; text-align: right;">Don't participate</td>
      <td style="padding: 16px 10px; background-color: #90EE90; font-weight: bold;">Don't pay cost; don't receive benefit</td>
      <td style="padding: 16px 10px; border-bottom: 3.5px solid #DC143C; border-left: 3.5px solid #DC143C; border-right: 3.5px solid #DC143C; background-color: #FFF5F5;">Don't pay cost; don't receive benefit</td>
      <td style="padding: 16px 10px;">Don't pay cost; receive benefit</td>
    </tr>
  </tbody>
</table>

---

## Collective Action Theory

<table style="margin: 25px auto; font-size: 0.85em; border-collapse: collapse; text-align: center; width: 95%;">
  <thead>
    <tr style="border-bottom: 2px solid #ccc;">
      <th style="padding: 12px 10px; width: 22%;"></th>
      <th style="padding: 12px 10px; width: 26%;">Scenario 1<br><span style="font-size: 0.85em; font-weight: normal;">Fewer than $K-1$ participate</span></th>
      <th style="padding: 12px 10px; width: 26%;">Scenario 2<br><span style="font-size: 0.85em; font-weight: normal;">Exactly $K-1$ participate</span></th>
      <th style="padding: 12px 10px; width: 26%; border-top: 3.5px solid #DC143C; border-left: 3.5px solid #DC143C; border-right: 3.5px solid #DC143C; background-color: #FFF5F5;">Scenario 3<br><span style="font-size: 0.85em; font-weight: normal;">$K$ or more participate</span></th>
    </tr>
  </thead>
  <tbody>
    <tr style="border-bottom: 1px solid #eee;">
      <td style="padding: 16px 10px; font-weight: bold; text-align: right;">Participate</td>
      <td style="padding: 16px 10px;">Pay cost; don't receive benefit</td>
      <td style="padding: 16px 10px; background-color: #90EE90; font-weight: bold;">Pay cost; receive benefit</td>
      <td style="padding: 16px 10px; border-left: 3.5px solid #DC143C; border-right: 3.5px solid #DC143C; background-color: #FFF5F5;">Pay cost; receive benefit</td>
    </tr>
    <tr>
      <td style="padding: 16px 10px; font-weight: bold; text-align: right;">Don't participate</td>
      <td style="padding: 16px 10px; background-color: #90EE90; font-weight: bold;">Don't pay cost; don't receive benefit</td>
      <td style="padding: 16px 10px;">Don't pay cost; don't receive benefit</td>
      <td style="padding: 16px 10px; border-bottom: 3.5px solid #DC143C; border-left: 3.5px solid #DC143C; border-right: 3.5px solid #DC143C; background-color: #FFF5F5;">Don't pay cost; receive benefit</td>
    </tr>
  </tbody>
</table>

---

## Collective Action Theory

<table style="margin: 25px auto; font-size: 0.85em; border-collapse: collapse; text-align: center; width: 95%;">
  <thead>
    <tr style="border-bottom: 2px solid #ccc;">
      <th style="padding: 12px 10px; width: 22%;"></th>
      <th style="padding: 12px 10px; width: 26%;">Scenario 1<br><span style="font-size: 0.85em; font-weight: normal;">Fewer than $K-1$ participate</span></th>
      <th style="padding: 12px 10px; width: 26%;">Scenario 2<br><span style="font-size: 0.85em; font-weight: normal;">Exactly $K-1$ participate</span></th>
      <th style="padding: 12px 10px; width: 26%; border-top: 3.5px solid #DC143C; border-left: 3.5px solid #DC143C; border-right: 3.5px solid #DC143C; background-color: #FFF5F5;">Scenario 3<br><span style="font-size: 0.85em; font-weight: normal;">$K$ or more participate</span></th>
    </tr>
  </thead>
  <tbody>
    <tr style="border-bottom: 1px solid #eee;">
      <td style="padding: 16px 10px; font-weight: bold; text-align: right;">Participate</td>
      <td style="padding: 16px 10px;">Pay cost; don't receive benefit</td>
      <td style="padding: 16px 10px; background-color: #90EE90; font-weight: bold;">Pay cost; receive benefit</td>
      <td style="padding: 16px 10px; border-left: 3.5px solid #DC143C; border-right: 3.5px solid #DC143C; background-color: #FFF5F5;">Pay cost; receive benefit</td>
    </tr>
    <tr>
      <td style="padding: 16px 10px; font-weight: bold; text-align: right;">Don't participate</td>
      <td style="padding: 16px 10px; background-color: #90EE90; font-weight: bold;">Don't pay cost; don't receive benefit</td>
      <td style="padding: 16px 10px;">Don't pay cost; don't receive benefit</td>
      <td style="padding: 16px 10px; border-bottom: 3.5px solid #DC143C; border-left: 3.5px solid #DC143C; border-right: 3.5px solid #DC143C; background-color: #90EE90; font-weight: bold;">Don't pay cost; receive benefit</td>
    </tr>
  </tbody>
</table>

---

## Collective Action Theory

<table style="margin: 25px auto; font-size: 0.85em; border-collapse: collapse; text-align: center; width: 95%;">
  <thead>
    <tr style="border-bottom: 2px solid #ccc;">
      <th style="padding: 12px 10px; width: 22%;"></th>
      <th style="padding: 12px 10px; width: 26%;">Scenario 1<br><span style="font-size: 0.85em; font-weight: normal;">Fewer than $K-1$ participate</span></th>
      <th style="padding: 12px 10px; width: 26%;">Scenario 2<br><span style="font-size: 0.85em; font-weight: normal;">Exactly $K-1$ participate</span></th>
      <th style="padding: 12px 10px; width: 26%;">Scenario 3<br><span style="font-size: 0.85em; font-weight: normal;">$K$ or more participate</span></th>
    </tr>
  </thead>
  <tbody>
    <tr style="border-bottom: 1px solid #eee;">
      <td style="padding: 16px 10px; font-weight: bold; text-align: right;">Participate</td>
      <td style="padding: 16px 10px;">Pay cost; don't receive benefit</td>
      <td style="padding: 16px 10px; background-color: #90EE90; font-weight: bold;">Pay cost; receive benefit</td>
      <td style="padding: 16px 10px;">Pay cost; receive benefit</td>
    </tr>
    <tr>
      <td style="padding: 16px 10px; font-weight: bold; text-align: right;">Don't participate</td>
      <td style="padding: 16px 10px; background-color: #90EE90; font-weight: bold;">Don't pay cost; don't receive benefit</td>
      <td style="padding: 16px 10px;">Don't pay cost; don't receive benefit</td>
      <td style="padding: 16px 10px; background-color: #90EE90; font-weight: bold;">Don't pay cost; receive benefit</td>
    </tr>
  </tbody>
</table>



## Tipping Models: Society A

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 11
#| fig-height: 4.4
#| out-width: "950px"
#| out-height: "360px"

library(grid)

draw_tipping_society <- function(thresholds, revolting_indices = c()) {
  grid.newpage()
  pushViewport(viewport(width = 1, height = 1))
  
  n <- length(thresholds)
  xs <- seq(0.07, 0.93, length.out = n)
  
  for (i in 1:n) {
    x <- xs[i]
    y_card <- 0.44
    t_val <- thresholds[i]
    is_revolting <- i %in% revolting_indices
    
    # Colors (Color-blind friendly: unfilled white vs vibrant green fill)
    if (is_revolting) {
      card_fill  <- "#C8E6C9"    # light green fill
      card_col   <- "#1B5E20"    # dark green stroke
      badge_fill <- "#2E7D32"    # green filled circle
      badge_text <- "white"
      badge_col  <- "#1B5E20"
      torso_col  <- "#1B5E20"
      card_lwd   <- 2.5
    } else {
      card_fill  <- "white"      # unfilled white
      card_col   <- "#718096"    # neutral outline
      badge_fill <- "white"      # unfilled circle
      badge_text <- "black"
      badge_col  <- "#4A5568"
      torso_col  <- "#A0AEC0"
      card_lwd   <- 1.5
    }
    
    # 1. Threshold circle above individual
    grid.circle(x = x, y = 0.77, r = 0.034, 
                gp = gpar(fill = badge_fill, col = badge_col, lwd = 1.6))
    grid.text(as.character(t_val), x = x, y = 0.77, 
              gp = gpar(fontsize = 13, fontface = "bold", col = badge_text))
    
    # 2. Person Card
    grid.roundrect(x = x, y = y_card, width = 0.075, height = 0.32, r = unit(0.015, "npc"),
                   gp = gpar(fill = card_fill, col = card_col, lwd = card_lwd))
    
    # 3. Person Icon
    # Head
    grid.circle(x = x, y = y_card + 0.075, r = 0.022, 
                gp = gpar(fill = "navajowhite", col = "black", lwd = 1.2))
    # Torso / Body
    grid.polygon(x = c(x - 0.026, x + 0.026, x + 0.020, x - 0.020),
                 y = c(y_card - 0.065, y_card - 0.065, y_card + 0.045, y_card + 0.045),
                 gp = gpar(fill = torso_col, col = "black", lwd = 1.2))
    
    # 4. Person Label (Index)
    grid.text(paste0("P", i), x = x, y = y_card - 0.11, 
              gp = gpar(fontsize = 10, fontface = "bold", col = "black"))
    
    # 5. Status indicator below card
    if (is_revolting) {
      grid.text("Revolts!", x = x, y = 0.16, 
                gp = gpar(fontsize = 10.5, fontface = "bold", col = "#1B5E20"))
    }
  }
  
  popViewport()
}

draw_tipping_society(c(0, 2, 2, 3, 4, 5, 6, 7, 8, 10), revolting_indices = c())
```

---

## Tipping Models: Society A

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 11
#| fig-height: 4.4
#| out-width: "950px"
#| out-height: "360px"

draw_tipping_society(c(0, 2, 2, 3, 4, 5, 6, 7, 8, 10), revolting_indices = c(1))
```


- **Person 1** ($T = 0$) protests alone $\rightarrow$ protest size is **1**
- **Persons 2 & 3** require at least 2 people protesting $\implies$ they stay home
- **Result:** The cascade fails to start $\rightarrow$ only **1** person revolts!

---

## Tipping Models: Society A'

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 11
#| fig-height: 4.4
#| out-width: "950px"
#| out-height: "360px"

draw_tipping_society(c(0, 1, 2, 3, 4, 5, 6, 7, 8, 10), revolting_indices = c())
```

---

## Tipping Models: Society A'

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 11
#| fig-height: 4.4
#| out-width: "950px"
#| out-height: "360px"

draw_tipping_society(c(0, 1, 2, 3, 4, 5, 6, 7, 8, 10), revolting_indices = 1:9)
```


- **Person 1** ($T = 0$) revolts $\rightarrow$ **Person 2** ($T = 1$) joins $\rightarrow$ **Person 3** ($T = 2$) joins $\dots$
- A complete **revolutionary cascade** is triggered!
- **Result:** **9** out of 10 people revolt (only loyalist $P_{10}$ stays home)
- $\implies$ A shift in just **one** person's threshold ($P_2: 2 \rightarrow 1$) causes massive change!

---

## Tipping Models: Society B

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 11
#| fig-height: 4.4
#| out-width: "950px"
#| out-height: "360px"

draw_tipping_society(c(0, 2, 3, 3, 4, 5, 6, 7, 8, 10), revolting_indices = c())
```

---

## Tipping Models: Society B

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 11
#| fig-height: 4.4
#| out-width: "950px"
#| out-height: "360px"

draw_tipping_society(c(0, 2, 3, 3, 4, 5, 6, 7, 8, 10), revolting_indices = c(1))
```


- **Person 1** ($T = 0$) revolts alone $\rightarrow$ protest size is **1**
- **Person 2** ($T = 2$) requires at least 2 people $\implies$ stays home
- **Result:** No cascade $\rightarrow$ only **1** person revolts

---

## Tipping Models: Society B'

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 11
#| fig-height: 4.4
#| out-width: "950px"
#| out-height: "360px"

draw_tipping_society(c(0, 1, 3, 3, 4, 5, 6, 7, 8, 10), revolting_indices = c())
```

---

## Tipping Models: Society B'

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 11
#| fig-height: 4.4
#| out-width: "950px"
#| out-height: "360px"

draw_tipping_society(c(0, 1, 3, 3, 4, 5, 6, 7, 8, 10), revolting_indices = c(1, 2))
```


- **Person 1** ($T = 0$) revolts $\rightarrow$ **Person 2** ($T = 1$) joins $\rightarrow$ protest size is **2**
- **Persons 3 & 4** require at least 3 people $\implies$ they stay home
- **Result:** Cascade is blocked by the gap at $T = 2 \rightarrow$ only **2** people revolt!

---

## Tipping Models: Society C

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 11
#| fig-height: 4.4
#| out-width: "950px"
#| out-height: "360px"

draw_tipping_society(c(0, 2, 2, 2, 2, 2, 2, 2, 2, 10), revolting_indices = c())
```

---

## Tipping Models: Society C

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 11
#| fig-height: 4.4
#| out-width: "950px"
#| out-height: "360px"

draw_tipping_society(c(0, 2, 2, 2, 2, 2, 2, 2, 2, 10), revolting_indices = c(1))
```


- Almost everyone is very willing to revolt ($T = 2$ for 8 out of 10 people)
- But **Person 1** ($T = 0$) protests alone $\rightarrow$ protest size is only **1**
- Without a critical bridge ($T = 1$), none of the $T = 2$ individuals join!
- **Result:** Highly discontented society, yet only **1** person revolts!

---

## Tipping Models: Society C'

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 11
#| fig-height: 4.4
#| out-width: "950px"
#| out-height: "360px"

draw_tipping_society(c(0, 0, 2, 2, 2, 2, 2, 2, 2, 10), revolting_indices = c())
```

---

## Tipping Models: Society C'

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 11
#| fig-height: 4.4
#| out-width: "950px"
#| out-height: "360px"

draw_tipping_society(c(0, 0, 2, 2, 2, 2, 2, 2, 2, 10), revolting_indices = 1:9)
```


- **Person 1** ($T = 0$) and **Person 2** ($T = 0$) both revolt $\rightarrow$ protest size is **2**
- All 7 individuals with $T = 2$ ($P_3 \dots P_9$) see their threshold met and join **simultaneously**!
- **Result:** Sudden, explosive cascade $\rightarrow$ **9** out of 10 people revolt!
- $\implies$ A shift in just one person's threshold ($P_2: 2 \rightarrow 0$) instantly tips the entire society!


# Top-Down Transitions {.section}


## Transition Game: Soft-Liners vs. Opposition

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 11
#| fig-height: 4.2
#| out-width: "900px"
#| out-height: "320px"

library(grid)

draw_transition_game_tree <- function(hl_branches = c(), hl_outcome = NULL, 
                                       branch_hl_col = "#DC143C", outcome_hl_col = "#DC143C") {
  grid.newpage()
  pushViewport(viewport(width = 1, height = 1))

  # Helper: Draw a decision node
  draw_node <- function(x, y, player) {
    grid.circle(x = x, y = y, r = 0.018, gp = gpar(fill = "black", col = "black"))
    grid.text(player, x = x, y = y + 0.065, gp = gpar(fontsize = 13, fontface = "bold", col = "black"))
  }

  # Helper: Draw an outcome box
  draw_outcome <- function(x, y, title, subtitle = NULL, is_hl = FALSE) {
    w <- 0.22
    h <- ifelse(is.null(subtitle), 0.085, 0.105)
    border_col <- ifelse(is_hl, outcome_hl_col, "black")
    fill_col   <- ifelse(is_hl, if (outcome_hl_col == "#DC143C") "#FFF5F5" else "#F1F8E9", "white")
    text_col   <- ifelse(is_hl, outcome_hl_col, "black")
    lwd_val    <- ifelse(is_hl, 3, 1.8)
    
    grid.roundrect(x = x + w/2 + 0.01, y = y, width = w, height = h, r = unit(0.015, "npc"),
                   gp = gpar(fill = fill_col, col = border_col, lwd = lwd_val))
    if (is.null(subtitle)) {
      grid.text(title, x = x + w/2 + 0.01, y = y, 
                gp = gpar(fontsize = 10.5, fontface = "bold", col = text_col))
    } else {
      grid.text(title, x = x + w/2 + 0.01, y = y + 0.020, 
                gp = gpar(fontsize = 10.5, fontface = "bold", col = text_col))
      grid.text(subtitle, x = x + w/2 + 0.01, y = y - 0.020, 
                gp = gpar(fontsize = 9, fontface = "italic", col = text_col))
    }
  }

  # Helper: Draw branch line + text
  draw_branch <- function(x1, y1, x2, y2, label, lx, ly, branch_id) {
    is_hl <- branch_id %in% hl_branches
    col_val <- ifelse(is_hl, branch_hl_col, "black")
    lwd_val <- ifelse(is_hl, 3.5, 2)
    grid.lines(x = c(x1, x2), y = c(y1, y2), gp = gpar(col = col_val, lwd = lwd_val))
    grid.text(label, x = lx, y = ly, gp = gpar(fontsize = 11.5, fontface = "bold", col = col_val))
  }

  # 1. ROOT (Stage 1: Soft-Liners)
  draw_branch(0.10, 0.55, 0.32, 0.80, "Do nothing", 0.17, 0.72, "do_nothing")
  draw_outcome(0.32, 0.80, "Status Quo", is_hl = (identical(hl_outcome, "status_quo")))

  draw_branch(0.10, 0.55, 0.35, 0.35, "Open", 0.20, 0.42, "open")
  draw_node(0.10, 0.55, "Soft-Liners")

  # 2. STAGE 2 (Opposition)
  draw_branch(0.35, 0.35, 0.58, 0.58, "Enter", 0.43, 0.51, "enter")
  draw_outcome(0.58, 0.58, "Broadened Dictatorship", is_hl = (identical(hl_outcome, "broadened_dictatorship")))

  draw_branch(0.35, 0.35, 0.60, 0.18, "Organize", 0.43, 0.23, "organize")
  draw_node(0.35, 0.35, "Opposition")

  # 3. STAGE 3 (Soft-Liners)
  draw_branch(0.60, 0.18, 0.76, 0.34, "Repress", 0.66, 0.30, "repress")
  draw_outcome(0.76, 0.34, "Narrow Dictatorship", "or Insurgency", is_hl = (identical(hl_outcome, "narrow_dictatorship")))

  draw_branch(0.60, 0.18, 0.76, 0.05, "Democratize", 0.66, 0.08, "democratize")
  draw_outcome(0.76, 0.05, "Democratic Transition", is_hl = (identical(hl_outcome, "democratic_transition")))
  draw_node(0.60, 0.18, "Soft-Liners")

  popViewport()
}

draw_transition_game_tree()
```

---

## Transition Game: Weak Opposition

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 11
#| fig-height: 4.2
#| out-width: "900px"
#| out-height: "320px"

draw_transition_game_tree(hl_branches = c("repress"))
```

::: {.nonincremental}
- **Stage 3 (Soft-Liners):** Facing a weak opposition, soft-liners prefer to [**Repress**]{style="color: #DC143C; font-weight: bold;"} rather than democratize
:::

---

## Transition Game: Weak Opposition

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 11
#| fig-height: 4.2
#| out-width: "900px"
#| out-height: "320px"

draw_transition_game_tree(hl_branches = c("repress", "enter"))
```

::: {.nonincremental}
- **Stage 3 (Soft-Liners):** Soft-liners will **Repress**
- **Stage 2 (Opposition):** Anticipating repression, the opposition decides to [**Enter**]{style="color: #DC143C; font-weight: bold;"} rather than organize
:::

---

## Transition Game: Weak Opposition

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 11
#| fig-height: 4.2
#| out-width: "900px"
#| out-height: "320px"

draw_transition_game_tree(hl_branches = c("repress", "enter", "open"))
```

::: {.nonincremental}
- **Stage 3 (Soft-Liners):** Soft-liners will **Repress**
- **Stage 2 (Opposition):** Opposition will **Enter**
- **Stage 1 (Soft-Liners):** Knowing the opposition will enter, soft-liners choose to [**Open**]{style="color: #DC143C; font-weight: bold;"}
:::

---

## Transition Game: Weak Opposition

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 11
#| fig-height: 4.2
#| out-width: "900px"
#| out-height: "320px"

draw_transition_game_tree(hl_branches = c("repress", "enter", "open"), hl_outcome = "broadened_dictatorship")
```

::: {.nonincremental}
- **Equilibrium Path:** Soft-Liners **Open** $\rightarrow$ Opposition **Enters**
- **Outcome:** [**Broadened Dictatorship**]{style="color: #DC143C; font-weight: bold;"} (liberalization without democratization!)
:::

---

## Transition Game: Strong Opposition

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 11
#| fig-height: 4.2
#| out-width: "900px"
#| out-height: "320px"

draw_transition_game_tree(hl_branches = c("democratize"))
```

::: {.nonincremental}
- **Stage 3 (Soft-Liners):** Repression fails against strong opposition $\implies$ soft-liners must [**Democratize**]{style="color: #DC143C; font-weight: bold;"}
:::

---

## Transition Game: Strong Opposition

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 11
#| fig-height: 4.2
#| out-width: "900px"
#| out-height: "320px"

draw_transition_game_tree(hl_branches = c("democratize", "organize"))
```

::: {.nonincremental}
- **Stage 3 (Soft-Liners):** Soft-liners will **Democratize**
- **Stage 2 (Opposition):** Anticipating democratization, the opposition chooses to [**Organize**]{style="color: #DC143C; font-weight: bold;"}
:::

---

## Transition Game: Strong Opposition

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 11
#| fig-height: 4.2
#| out-width: "900px"
#| out-height: "320px"

draw_transition_game_tree(hl_branches = c("democratize", "organize", "do_nothing"))
```

::: {.nonincremental}
- **Stage 3 (Soft-Liners):** Soft-liners will **Democratize**
- **Stage 2 (Opposition):** Opposition will **Organize**
- **Stage 1 (Soft-Liners):** Knowing opening triggers democratization, soft-liners choose to [**Do nothing**]{style="color: #DC143C; font-weight: bold;"}
:::

---

## Transition Game: Strong Opposition

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 11
#| fig-height: 4.2
#| out-width: "900px"
#| out-height: "320px"

draw_transition_game_tree(hl_branches = c("democratize", "organize", "do_nothing"), hl_outcome = "status_quo")
```

::: {.nonincremental}
- **Equilibrium Path:** Soft-Liners **Do nothing**
- **Outcome:** [**Status Quo**]{style="color: #DC143C; font-weight: bold;"} (under full information, democratization never happens!)
:::

---

## Democratic Transition by Mistake

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 11
#| fig-height: 4.2
#| out-width: "900px"
#| out-height: "320px"

draw_transition_game_tree(hl_branches = c("open"), branch_hl_col = "#2E7D32")
```

::: {.nonincremental}
- **Belief:** Soft-liners mistakenly believe the opposition is **weak** (expecting them to *Enter*)
- **Stage 1 (Soft-Liners):** Based on this miscalculation, soft-liners choose to [**Open**]{style="color: #2E7D32; font-weight: bold;"}
:::

---

## Democratic Transition by Mistake

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 11
#| fig-height: 4.2
#| out-width: "900px"
#| out-height: "320px"

draw_transition_game_tree(hl_branches = c("open", "organize"), branch_hl_col = "#2E7D32")
```

::: {.nonincremental}
- **Stage 1:** Soft-liners **Open** (expecting opposition to enter)
- **Stage 2 (Opposition):** But the opposition is actually **strong** and chooses to [**Organize**]{style="color: #2E7D32; font-weight: bold;"}!
:::

---

## Democratic Transition by Mistake

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 11
#| fig-height: 4.2
#| out-width: "900px"
#| out-height: "320px"

draw_transition_game_tree(hl_branches = c("open", "organize", "democratize"), 
                          hl_outcome = "democratic_transition",
                          branch_hl_col = "#2E7D32", outcome_hl_col = "#DC143C")
```

::: {.nonincremental}
- **Stage 1:** Soft-liners **Open** (mistaken belief that opposition is weak)
- **Stage 2:** Opposition **Organizes** (revealing true strength)
- **Stage 3 (Soft-Liners):** Mistake realized — repression would fail $\implies$ soft-liners must [**Democratize**]{style="color: #2E7D32; font-weight: bold;"}
- **Outcome:** [**Democratic Transition**]{style="color: #DC143C; font-weight: bold;"} occurs **by mistake** (miscalculation / incomplete information)!
:::

# Summing Up {.section}

## 

::: {style="display: flex; justify-content: center; align-items: center; height: 70vh; font-size: 2.5em; font-weight: bold; color: Black;"}
Thank you!
:::