---
title: "6. Varieties of Dictatorship"
subtitle: "Comparative Politics"
format: revealjs
---


## Some Myths about Dictators

- We tend to imagine dictators as omnipotent individuals

- They actually govern through complex institutions

## Our Plan for Today

Types of dictatorship

- Does it matter if the dictator is a king or a general?

. . . 

How Dictators rule

:::{.nonincremental}
- Why do dictatorship have parties and elections?
- Media and propaganda
- Violence and repression
:::

. . . 

Explaining dictators' economic performance

:::{.nonincremental}
- [Selectorate]{.alert} theory
:::


# Types of Dictatorship {.section}


## Monarchic Dictatorships

[Figure TBA]

- The dictator is a king
- Key power roles are held within the royal family
- Succession happens within the royal family

Tend to be more stable than other regimes. Why?

## Monarchic Dictatorships

Tend to be more stable than other regimes. Why?

:::: {.columns}

::: {.column width="48%"}

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 5.5
#| fig-height: 5.2
#| out-width: "100%"

library(grid)

draw_monarchic_network <- function() {
  grid.newpage()
  pushViewport(viewport(width = 1, height = 1))
  
  # 7 Nodes (King is node 1: embedded in the network, not the sole center)
  nodes <- data.frame(
    id = 1:7,
    x = c(0.72, 0.36, 0.16, 0.28, 0.62, 0.85, 0.48),
    y = c(0.70, 0.80, 0.50, 0.24, 0.20, 0.45, 0.50),
    is_king = c(TRUE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE)
  )
  
  # Dense kinship and power-sharing ties
  edges <- matrix(c(
    1, 2,  1, 6,  1, 7,  1, 5,
    2, 3,  2, 7,  2, 4,
    3, 4,  3, 7,
    4, 5,  4, 7,
    5, 6,  5, 7,
    6, 7
  ), ncol = 2, byrow = TRUE)
  
  # 1. Background boundary indicating the "Ruling Family"
  hull_idx <- c(3, 2, 1, 6, 5, 4)
  hx <- nodes$x[hull_idx]
  hy <- nodes$y[hull_idx]
  cx <- mean(nodes$x)
  cy <- mean(nodes$y)
  hx_exp <- cx + (hx - cx) * 1.30
  hy_exp <- cy + (hy - cy) * 1.28
  
  grid.polygon(hx_exp, hy_exp, 
               gp = gpar(fill = "#F8FAFC", col = "#CBD5E1", lwd = 1.8, lty = "dashed"))
  
  # Title above network
  grid.text("Royal Family Network", x = 0.50, y = 0.94, 
            gp = gpar(fontsize = 14, fontface = "bold", col = "#0F172A"))
  
  # 2. Draw Edges
  for (e in 1:nrow(edges)) {
    n1 <- edges[e, 1]
    n2 <- edges[e, 2]
    grid.lines(x = c(nodes$x[n1], nodes$x[n2]), 
               y = c(nodes$y[n1], nodes$y[n2]), 
               gp = gpar(col = "#94A3B8", lwd = 2.0))
  }
  
  # Helper for Crown
  draw_crown <- function(x, y, w, h) {
    xs <- x + c(-w/2, -w/2, -w/4, 0, w/4, w/2, w/2)
    ys <- y + c(0, h*0.7, h*0.3, h, h*0.3, h*0.7, 0)
    grid.polygon(xs, ys, gp = gpar(fill = "#F59E0B", col = "#B45309", lwd = 1.4))
    
    # Jewels
    tip_xs <- x + c(-w/2, 0, w/2)
    tip_ys <- y + c(h*0.7, h, h*0.7)
    for (i in 1:3) {
      grid.circle(tip_xs[i], tip_ys[i] + h*0.09, r = h*0.12, 
                  gp = gpar(fill = "#EF4444", col = "#991B1B", lwd = 0.8))
    }
  }
  
  # 3. Draw Nodes
  r_node <- 0.056
  for (i in 1:nrow(nodes)) {
    nx <- nodes$x[i]
    ny <- nodes$y[i]
    is_k <- nodes$is_king[i]
    
    border_col <- if (is_k) "#D97706" else "#475569"
    fill_col   <- if (is_k) "#FEF3C7" else "#FFFFFF"
    lwd_val    <- if (is_k) 2.6 else 1.8
    
    grid.circle(nx, ny, r = r_node, 
                gp = gpar(fill = fill_col, col = border_col, lwd = lwd_val))
    
    head_r <- r_node * 0.35
    head_y <- ny + r_node * 0.16
    grid.circle(nx, head_y, r = head_r, 
                gp = gpar(fill = "#FDE68A", col = "#1F2937", lwd = 1.1))
    
    torso_col <- if (is_k) "#DC2626" else "#94A3B8"
    tx <- c(nx - r_node*0.54, nx + r_node*0.54, nx + r_node*0.40, nx - r_node*0.40)
    ty <- c(ny - r_node*0.82, ny - r_node*0.82, ny - r_node*0.06, ny - r_node*0.06)
    grid.polygon(tx, ty, gp = gpar(fill = torso_col, col = "#1F2937", lwd = 1.1))
    
    if (is_k) {
      draw_crown(nx, head_y + head_r * 0.82, w = head_r * 2.2, h = head_r * 1.5)
    }
  }
  
  popViewport()
}

draw_monarchic_network()
```

:::

::: {.column width="52%"}

- Everyone has an interest in government stability

- Power sharing and succession encoded in family norms 

- Family members monitor the king $\implies$ [commitment]{.alert}

. . . 

$\implies$ Less incentives from powerful elites to overthrow the king

:::

::::



## Military Dictatorships

[Figure TBA]

- The dictator is/was a member of the military
- Often come to power through military coups
- Motivated by threat of unrest or economic interests

Relatively less stable. Why?


## Military Dictatorships

Relatively less stable. Why?

:::: {.columns}

::: {.column width="48%"}

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 5.5
#| fig-height: 5.2
#| out-width: "100%"

library(grid)

draw_military_networks <- function() {
  grid.newpage()
  pushViewport(viewport(width = 1, height = 1))
  
  # Title
  grid.text("Military Junta and Factions", x = 0.50, y = 0.94, 
            gp = gpar(fontsize = 14, fontface = "bold", col = "#0F172A"))
  
  # Helper to draw a military person node
  draw_military_node <- function(x, y, r = 0.052, torso_col = "#2D5A27", is_junta = FALSE) {
    fill_col <- if (is_junta) "#F0FDF4" else "#FFFFFF"
    b_col    <- if (is_junta) "#15803D" else "#475569"
    lwd_val  <- if (is_junta) 2.4 else 1.8
    
    grid.circle(x, y, r = r, 
                gp = gpar(fill = fill_col, col = b_col, lwd = lwd_val))
    
    head_r <- r * 0.35
    head_y <- y + r * 0.16
    grid.circle(x, head_y, r = head_r, 
                gp = gpar(fill = "#FDE68A", col = "#1F2937", lwd = 1.1))
    
    tx <- c(x - r*0.54, x + r*0.54, x + r*0.40, x - r*0.40)
    ty <- c(y - r*0.82, y - r*0.82, y - r*0.06, y - r*0.06)
    grid.polygon(tx, ty, gp = gpar(fill = torso_col, col = "#1F2937", lwd = 1.1))
    
    grid.circle(x, y - r*0.38, r = r*0.10, 
                gp = gpar(fill = "#F59E0B", col = "#B45309", lwd = 0.7))
  }
  
  # Helper for dashed boundary box
  draw_cluster_boundary <- function(x, y, w, h, label = NULL, col = "#CBD5E1", fill = "#F8FAFC") {
    grid.roundrect(x, y, width = w, height = h, r = unit(0.04, "npc"),
                   gp = gpar(fill = fill, col = col, lwd = 1.6, lty = "dashed"))
    if (!is.null(label)) {
      grid.text(label, x = x, y = y + h/2 - 0.042, 
                gp = gpar(fontsize = 11, fontface = "bold", col = "#1E293B"))
    }
  }
  
  # 1. CLUSTER BOUNDARIES
  # A. Military Junta (Top)
  draw_cluster_boundary(0.50, 0.67, w = 0.76, h = 0.43, 
                        label = "Military Junta", col = "#86EFAC", fill = "#F0FDF4")
  
  # B. Faction A (Bottom Left)
  draw_cluster_boundary(0.26, 0.22, w = 0.44, h = 0.36, 
                        label = "Faction A", col = "#CBD5E1", fill = "#F8FAFC")
  
  # C. Faction B (Bottom Right)
  draw_cluster_boundary(0.74, 0.22, w = 0.44, h = 0.36, 
                        label = "Faction B", col = "#CBD5E1", fill = "#F8FAFC")
  
  # 2. EDGES & NODES
  # Military Junta (4 nodes)
  j_nodes <- data.frame(
    x = c(0.34, 0.66, 0.50, 0.50),
    y = c(0.68, 0.68, 0.77, 0.53)
  )
  j_edges <- matrix(c(
    1, 2,  1, 3,  1, 4,
    2, 3,  2, 4,
    3, 4
  ), ncol = 2, byrow = TRUE)
  
  for (e in 1:nrow(j_edges)) {
    n1 <- j_edges[e, 1]
    n2 <- j_edges[e, 2]
    grid.lines(x = c(j_nodes$x[n1], j_nodes$x[n2]), 
               y = c(j_nodes$y[n1], j_nodes$y[n2]), 
               gp = gpar(col = "#94A3B8", lwd = 2.0))
  }
  for (i in 1:nrow(j_nodes)) {
    draw_military_node(j_nodes$x[i], j_nodes$y[i], r = 0.052, 
                       torso_col = "#1E4D2B", is_junta = TRUE)
  }
  
  # Faction A (3 nodes)
  f1_nodes <- data.frame(
    x = c(0.18, 0.34, 0.26),
    y = c(0.24, 0.24, 0.11)
  )
  f1_edges <- matrix(c(1, 2, 2, 3, 3, 1), ncol = 2, byrow = TRUE)
  
  for (e in 1:nrow(f1_edges)) {
    n1 <- f1_edges[e, 1]
    n2 <- f1_edges[e, 2]
    grid.lines(x = c(f1_nodes$x[n1], f1_nodes$x[n2]), 
               y = c(f1_nodes$y[n1], f1_nodes$y[n2]), 
               gp = gpar(col = "#94A3B8", lwd = 2.0))
  }
  for (i in 1:nrow(f1_nodes)) {
    draw_military_node(f1_nodes$x[i], f1_nodes$y[i], r = 0.046, 
                       torso_col = "#3B6E44", is_junta = FALSE)
  }
  
  # Faction B (2 nodes)
  f2_nodes <- data.frame(
    x = c(0.66, 0.82),
    y = c(0.18, 0.18)
  )
  grid.lines(x = c(f2_nodes$x[1], f2_nodes$x[2]), 
             y = c(f2_nodes$y[1], f2_nodes$y[2]), 
             gp = gpar(col = "#94A3B8", lwd = 2.0))
  
  for (i in 1:nrow(f2_nodes)) {
    draw_military_node(f2_nodes$x[i], f2_nodes$y[i], r = 0.046, 
                       torso_col = "#3B6E44", is_junta = FALSE)
  }
  
  popViewport()
}

draw_military_networks()
```

:::

::: {.column width="52%"}

- Armed forces easily split into factions

- Disagreement over everyday politics leads to conflict

- Diffused military strength makes balance of power uncertain

- Military may prefer to "allow" democracy rather than maintaining power

:::

::::



## Civilian Dictatorships

:::: {.columns}

::: {.column width="48%"}

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 5.5
#| fig-height: 5.2
#| out-width: "100%"

library(grid)

draw_civilian_dictatorship <- function() {
  grid.newpage()
  pushViewport(viewport(width = 1, height = 1))
  
  # Title
  grid.text("Civilian Dictatorship", x = 0.50, y = 0.94, 
            gp = gpar(fontsize = 14, fontface = "bold", col = "#0F172A"))
  
  # Helper to draw a stylized civilian node
  draw_civilian_node <- function(x, y, r = 0.050, torso_col = "#64748B", is_dictator = FALSE) {
    fill_col <- if (is_dictator) "#EFF6FF" else "#FFFFFF"
    b_col    <- if (is_dictator) "#1E3A8A" else "#475569"
    lwd_val  <- if (is_dictator) 2.6 else 1.8
    
    # Outer circle
    grid.circle(x, y, r = r, 
                gp = gpar(fill = fill_col, col = b_col, lwd = lwd_val))
    
    # Head
    head_r <- r * 0.35
    head_y <- y + r * 0.16
    grid.circle(x, head_y, r = head_r, 
                gp = gpar(fill = "#FDE68A", col = "#1F2937", lwd = 1.1))
    
    # Torso (civilian clothing)
    tx <- c(x - r*0.54, x + r*0.54, x + r*0.40, x - r*0.40)
    ty <- c(y - r*0.82, y - r*0.82, y - r*0.06, y - r*0.06)
    grid.polygon(tx, ty, gp = gpar(fill = torso_col, col = "#1F2937", lwd = 1.1))
    
    # Dictator detail: dark suit with red necktie
    if (is_dictator) {
      grid.polygon(x = c(x, x - r*0.13, x, x + r*0.13),
                   y = c(y - r*0.06, y - r*0.45, y - r*0.68, y - r*0.45),
                   gp = gpar(fill = "#DC2626", col = "#991B1B", lwd = 0.8))
    }
  }
  
  # 1. Dictator at the top
  x_dict <- 0.50
  y_dict <- 0.78
  r_dict <- 0.058
  
  # 2. Nodes below
  # Group 1 (Organized network, e.g. party/institutions): 3 nodes
  g1_nodes <- data.frame(
    x = c(0.18, 0.36, 0.27),
    y = c(0.40, 0.40, 0.22)
  )
  g1_edges <- matrix(c(1, 2, 2, 3, 3, 1), ncol = 2, byrow = TRUE)
  
  # Group 2 (Organized network, e.g. elites/interest groups): 2 nodes
  g2_nodes <- data.frame(
    x = c(0.68, 0.84),
    y = c(0.38, 0.24)
  )
  
  # Standalone unconnected individuals
  ind_nodes <- data.frame(
    x = c(0.50, 0.88, 0.14),
    y = c(0.18, 0.48, 0.16)
  )
  
  # Internal network edges (grey)
  for (e in 1:nrow(g1_edges)) {
    n1 <- g1_edges[e, 1]
    n2 <- g1_edges[e, 2]
    grid.lines(x = c(g1_nodes$x[n1], g1_nodes$x[n2]), 
               y = c(g1_nodes$y[n1], g1_nodes$y[n2]), 
               gp = gpar(col = "#94A3B8", lwd = 2.0))
  }
  
  grid.lines(x = c(g2_nodes$x[1], g2_nodes$x[2]), 
             y = c(g2_nodes$y[1], g2_nodes$y[2]), 
             gp = gpar(col = "#94A3B8", lwd = 2.0))
  
  # 3. Two-sided arrows (Dictator <-> Groups & Citizens)
  # Arrow 1: Dictator <-> Group 1 (left)
  grid.lines(x = c(x_dict - 0.055, 0.31), 
             y = c(y_dict - 0.055, 0.46), 
             arrow = arrow(ends = "both", type = "closed", length = unit(0.024, "npc")),
             gp = gpar(col = "#2563EB", fill = "#2563EB", lwd = 2.2))
  
  # Arrow 2: Dictator <-> Group 2 (right)
  grid.lines(x = c(x_dict + 0.055, 0.70), 
             y = c(y_dict - 0.055, 0.45), 
             arrow = arrow(ends = "both", type = "closed", length = unit(0.024, "npc")),
             gp = gpar(col = "#2563EB", fill = "#2563EB", lwd = 2.2))
  
  # Arrow 3: Dictator <-> Unconnected citizens (center bottom)
  grid.lines(x = c(x_dict, 0.50), 
             y = c(y_dict - 0.075, 0.27), 
             arrow = arrow(ends = "both", type = "closed", length = unit(0.024, "npc")),
             gp = gpar(col = "#2563EB", fill = "#2563EB", lwd = 2.2))
  
  # 4. Draw Nodes
  for (i in 1:nrow(g1_nodes)) {
    draw_civilian_node(g1_nodes$x[i], g1_nodes$y[i], r = 0.048, torso_col = "#475569")
  }
  for (i in 1:nrow(g2_nodes)) {
    draw_civilian_node(g2_nodes$x[i], g2_nodes$y[i], r = 0.048, torso_col = "#475569")
  }
  for (i in 1:nrow(ind_nodes)) {
    draw_civilian_node(ind_nodes$x[i], ind_nodes$y[i], r = 0.045, torso_col = "#64748B")
  }
  
  # Dictator node
  draw_civilian_node(x_dict, y_dict, r = r_dict, torso_col = "#1E3A8A", is_dictator = TRUE)
  
  popViewport()
}

draw_civilian_dictatorship()
```

:::

::: {.column width="52%"}

- The dictator is a civilian (not a member of the army)

- Needs to build support base out of existing institutions

- Use parties or personality cults

:::

::::


## Civilian Dictatorships: Dominant Party

- Parties as tools of social mobilization (rather than representation)
- Career in the party = career in the regime

## Civilian Dictatorships: Personalist

## Types of Dictatorships in Time

[figure from book]

## Classifying Dictatorships

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 10
#| fig-height: 5.5
#| out-width: "950px"
#| out-height: "520px"

library(grid)

grid.newpage()
pushViewport(viewport(width = 1, height = 1))

col_line <- "#111827"
col_text <- "#111827"
col_source <- "#4B5563"
lwd_line <- 3.2

# Coordinates
y_q1 <- 0.94
y_q2 <- 0.79
y_b1 <- 0.67       # split 1
y_yesno1 <- 0.58   # Yes/No 1
y_monarchy <- 0.44 # MONARCHY

y_q3 <- 0.45       # Q3
y_b2 <- 0.31       # split 2
y_yesno2 <- 0.22   # Yes/No 2
y_mil_civ <- 0.09  # MILITARY / CIVILIAN

x_center <- 0.50
x_left1 <- 0.28    # Monarchy branch
x_right1 <- 0.72   # Q3 branch

x_left2 <- 0.56    # Military branch
x_right2 <- 0.88   # Civilian branch

# 1. Question 1
grid.text("1. Who is the effective head of government?", 
          x = x_center, y = y_q1, 
          gp = gpar(fontsize = 15, fontface = "bold", col = col_text))

# Line Q1 -> Q2
grid.lines(x = c(x_center, x_center), y = c(y_q1 - 0.032, y_q2 + 0.052), 
           gp = gpar(col = col_line, lwd = lwd_line, lineend = "square"))

# 2. Question 2
grid.text("2. Does the effective head of government bear the title of \"king\"", 
          x = x_center, y = y_q2 + 0.022, 
          gp = gpar(fontsize = 14, fontface = "bold", col = col_text))
grid.text("and have a hereditary successor or predecessor?", 
          x = x_center, y = y_q2 - 0.022, 
          gp = gpar(fontsize = 14, fontface = "bold", col = col_text))

# Split 1
grid.lines(x = c(x_center, x_center), y = c(y_q2 - 0.052, y_b1), 
           gp = gpar(col = col_line, lwd = lwd_line, lineend = "square"))
grid.lines(x = c(x_left1, x_right1), y = c(y_b1, y_b1), 
           gp = gpar(col = col_line, lwd = lwd_line, lineend = "square"))
grid.lines(x = c(x_left1, x_left1), y = c(y_b1, y_yesno1 + 0.03), 
           gp = gpar(col = col_line, lwd = lwd_line, lineend = "square"))
grid.lines(x = c(x_right1, x_right1), y = c(y_b1, y_yesno1 + 0.03), 
           gp = gpar(col = col_line, lwd = lwd_line, lineend = "square"))

# Labels (Split 1)
grid.text("Yes", x = x_left1, y = y_yesno1, 
          gp = gpar(fontsize = 14, fontface = "plain", col = col_text))
grid.text("No", x = x_right1, y = y_yesno1, 
          gp = gpar(fontsize = 14, fontface = "plain", col = col_text))

# Yes -> MONARCHY
grid.lines(x = c(x_left1, x_left1), y = c(y_yesno1 - 0.03, y_monarchy + 0.04), 
           gp = gpar(col = col_line, lwd = lwd_line, lineend = "square"))
grid.text("MONARCHY", x = x_left1, y = y_monarchy, 
          gp = gpar(fontsize = 15, fontface = "bold", col = col_text))

# No -> Q3
grid.lines(x = c(x_right1, x_right1), y = c(y_yesno1 - 0.03, y_q3 + 0.052), 
           gp = gpar(col = col_line, lwd = lwd_line, lineend = "square"))

# 3. Question 3
grid.text("3. Is the effective head of government a current or", 
          x = x_right1, y = y_q3 + 0.022, 
          gp = gpar(fontsize = 13.5, fontface = "bold", col = col_text))
grid.text("past member of the armed forces?", 
          x = x_right1, y = y_q3 - 0.022, 
          gp = gpar(fontsize = 13.5, fontface = "bold", col = col_text))

# Split 2
grid.lines(x = c(x_right1, x_right1), y = c(y_q3 - 0.052, y_b2), 
           gp = gpar(col = col_line, lwd = lwd_line, lineend = "square"))
grid.lines(x = c(x_left2, x_right2), y = c(y_b2, y_b2), 
           gp = gpar(col = col_line, lwd = lwd_line, lineend = "square"))
grid.lines(x = c(x_left2, x_left2), y = c(y_b2, y_yesno2 + 0.03), 
           gp = gpar(col = col_line, lwd = lwd_line, lineend = "square"))
grid.lines(x = c(x_right2, x_right2), y = c(y_b2, y_yesno2 + 0.03), 
           gp = gpar(col = col_line, lwd = lwd_line, lineend = "square"))

# Labels (Split 2)
grid.text("Yes", x = x_left2, y = y_yesno2, 
          gp = gpar(fontsize = 14, fontface = "plain", col = col_text))
grid.text("No", x = x_right2, y = y_yesno2, 
          gp = gpar(fontsize = 14, fontface = "plain", col = col_text))

# Lines to outcomes
grid.lines(x = c(x_left2, x_left2), y = c(y_yesno2 - 0.03, y_mil_civ + 0.04), 
           gp = gpar(col = col_line, lwd = lwd_line, lineend = "square"))
grid.lines(x = c(x_right2, x_right2), y = c(y_yesno2 - 0.03, y_mil_civ + 0.04), 
           gp = gpar(col = col_line, lwd = lwd_line, lineend = "square"))

grid.text("MILITARY", x = x_left2, y = y_mil_civ, 
          gp = gpar(fontsize = 15, fontface = "bold", col = col_text))
grid.text("CIVILIAN", x = x_right2, y = y_mil_civ, 
          gp = gpar(fontsize = 15, fontface = "bold", col = col_text))

# Footnote
grid.text("(Adapted from Cheibub, Gandhi, and Vreeland (2010) and from Clark, Golder, and Golder)", 
          x = 0.03, y = 0.015, just = "left", 
          gp = gpar(fontsize = 10, fontface = "italic", col = col_source))

popViewport()
```



# Authoritarian Government {.section}

## Competitive Authoritarianism

## Personality Cults

[use to talk about propaganda]

## Spin Dictators


# Fundamental Problems of Authoritarian Rule {.section}